---
title: "SHANK3 ASO Mechanism - Intermediate Final Report"
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
root <- if (file.exists("_config.yml")) getwd() else if (file.exists("../_config.yml")) normalizePath("..") else getwd()
hyp <- readr::read_tsv(file.path(root, "results/07_mechanism_hypotheses.tsv"), show_col_types = FALSE)
sumtxt <- tryCatch(readLines(file.path(root, "results/PIPELINE_SUMMARY.txt")), error = function(e) "Summary not yet written.")
```

# Chemistry reminder

Core chemistry **MOE** (all bases in MOE brackets), PS backbone, architecture **likely_uniform_MOE**, gapmer **not supported**, RNase-H(SHANK3) **low_support**, 5m-dC **reported_but_needs_confirmation**, FAM present but excluded from unlabeled mechanism. Phenotype: increased SHANK3 expression for both ASOs.

# Top mechanism hypotheses

```{r}
knitr::kable(hyp %>% dplyr::filter(rank <= 3))
```

# Pipeline summary

```{r, results='asis'}
cat(paste(sumtxt, collapse = "\n"))
```

See `FINAL_SHANK3_ASO_MECHANISM_REPORT.Rmd` for the full 15-section report.
