---
title: "FINAL SHANK3 ASO Mechanism-of-Action Report"
author: "R / targets chemistry-aware bioinformatics workflow"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: flatly
    number_sections: true
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
    keep_tex: false
    fig_crop: false
    includes:
      in_header: final_report_header.tex
    pandoc_args:
      - "--variable=geometry:margin=0.65in"
mainfont: DejaVu Sans
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, results = "asis",
                      fig.width = 8, fig.height = 4)
suppressPackageStartupMessages({
  library(dplyr)
  library(readr)
  library(yaml)
  library(knitr)
})
has_kableExtra <- requireNamespace("kableExtra", quietly = TRUE)
if (has_kableExtra) suppressPackageStartupMessages(library(kableExtra))

root <- if (file.exists("_config.yml")) {
  getwd()
} else if (file.exists("../_config.yml")) {
  normalizePath("..")
} else {
  getwd()
}
cfg <- yaml::read_yaml(file.path(root, "_config.yml"))
read_safe <- function(path) {
  if (!file.exists(path)) return(NULL)
  tryCatch(readr::read_tsv(path, show_col_types = FALSE), error = function(e) NULL)
}
qc <- read_safe(file.path(root, "results/01_sequence_qc.tsv"))
map <- read_safe(file.path(root, "results/02_shank3_mapping.tsv"))
struct <- read_safe(file.path(root, "results/03_rna_structure.tsv"))
clip <- read_safe(file.path(root, "results/04_rbp_clip_overlap.tsv"))
motif <- read_safe(file.path(root, "results/04_rbp_motif_overlap.tsv"))
expr <- read_safe(file.path(root, "results/04_rbp_expression.tsv"))
func <- read_safe(file.path(root, "results/04_rbp_function.tsv"))
tf <- read_safe(file.path(root, "results/05_tf_chromatin.tsv"))
ranked <- read_safe(file.path(root, "results/06_ranked_rbps.tsv"))
hyp <- read_safe(file.path(root, "results/07_mechanism_hypotheses.tsv"))
val <- read_safe(file.path(root, "results/07_validation_recommendations.tsv"))
prov <- read_safe(file.path(root, "data/raw/PROVENANCE.tsv"))
`%||%` <- function(a, b) if (is.null(a) || length(a) == 0 || (is.character(a) && !nzchar(a[[1]]))) b else a

is_pdf <- knitr::is_latex_output()

shorten <- function(x, n = if (is_pdf) 32 else 120) {
  x <- as.character(x)
  ifelse(is.na(x), NA_character_,
         ifelse(nchar(x) > n, paste0(substr(x, 1, max(1, n - 3)), "..."), x))
}

fmt_num <- function(x) {
  ifelse(
    is.na(x), NA_character_,
    ifelse(abs(x - round(x)) < 1e-8,
           format(as.integer(round(x)), scientific = FALSE, trim = TRUE, big.mark = ""),
           format(round(x, 2), nsmall = 1, scientific = FALSE, trim = TRUE))
  )
}

fmt_tbl <- function(df, max_rows = NULL) {
  if (is.null(df) || !nrow(df)) return(df)
  if (!is.null(max_rows) && nrow(df) > max_rows) df <- utils::head(df, max_rows)
  df <- as.data.frame(df, stringsAsFactors = FALSE)
  for (nm in names(df)) {
    if (is.character(df[[nm]]) || is.factor(df[[nm]])) {
      df[[nm]] <- shorten(as.character(df[[nm]]))
    } else if (is.logical(df[[nm]])) {
      df[[nm]] <- ifelse(is.na(df[[nm]]), "NA", ifelse(df[[nm]], "T", "F"))
    } else if (is.numeric(df[[nm]])) {
      df[[nm]] <- fmt_num(df[[nm]])
    }
  }
  df
}

# PDF-safe tables: never longtable+resizebox; scale_down tabular; optional landscape.
show_kable <- function(df, caption = NULL, max_rows = NULL, font_size = 9,
                       landscape = FALSE) {
  if (is.null(df) || !nrow(df)) {
    cat("*No rows to display.*\n\n")
    return(invisible(NULL))
  }
  df2 <- fmt_tbl(df, max_rows = max_rows)
  if (!is_pdf) {
    cat(knitr::kable(df2, format = "markdown", caption = caption), sep = "\n")
    cat("\n\n")
    return(invisible(NULL))
  }

  max_cols <- if (landscape) 8 else 5
  if (ncol(df2) > max_cols) df2 <- df2[, seq_len(max_cols), drop = FALSE]

  emit_one <- function(piece, cap = NULL) {
    kb <- knitr::kable(
      piece,
      format = "latex",
      booktabs = TRUE,
      longtable = FALSE,
      caption = cap,
      align = paste(rep("l", ncol(piece)), collapse = ""),
      escape = TRUE,
      row.names = FALSE
    )
    if (has_kableExtra) {
      kb <- kableExtra::kable_styling(
        kb,
        latex_options = c("hold_position", "scale_down"),
        font_size = font_size,
        full_width = FALSE
      )
      if (isTRUE(landscape)) kb <- kableExtra::landscape(kb)
    }
    cat(as.character(kb), sep = "\n")
    cat("\n\n")
  }

  chunk_n <- if (landscape) 16 else 12
  if (nrow(df2) <= chunk_n) {
    emit_one(df2, caption)
  } else {
    starts <- seq(1, nrow(df2), by = chunk_n)
    for (i in seq_along(starts)) {
      s <- starts[[i]]
      e <- min(nrow(df2), s + chunk_n - 1)
      cap <- if (i == 1) caption else paste0(caption %||% "Table", " (continued)")
      emit_one(df2[s:e, , drop = FALSE], cap)
    }
  }
  invisible(NULL)
}

tsv_note <- function(rel_path) {
  cat("\n*Full table (all columns):* `", rel_path, "`\n\n", sep = "")
  invisible(NULL)
}
```

# Executive summary

This report summarizes a reproducible **R-only** workflow for two SHANK3-associated ASOs
(`r paste(vapply(cfg$asos, function(x) x$id, character(1)), collapse = ", ")`)
in *Homo sapiens* **GRCh38**. Both ASOs are linked to **increased SHANK3 expression**. Chemistry (revised): **core=MOE**, all bases in MOE brackets, **PS backbone**, architecture **likely_uniform_MOE**, gapmer **not supported** by current supplier notation, RNase-H direct SHANK3 mRNA **low_support**, 5m-dC **reported but needs confirmation**, FAM present in supplier construct but **excluded** from unlabeled mechanism interpretation. Sequence/database evidence is used only to generate **ranked hypotheses**, not confirmed mechanisms.

```{r}
exact_n <- if (is.null(map)) 0 else sum(map$exact_shank3_match, na.rm = TRUE)
approx_n <- if (is.null(map)) 0 else sum(map$approximate_shank3_match, na.rm = TRUE)
top_mech <- if (is.null(hyp)) NULL else hyp %>% dplyr::filter(is_top)
show_kable(data.frame(
  metric = c("Exact SHANK3 mapping rows", "Approximate mapping rows", "Top mechanism labels"),
  value = c(
    as.character(exact_n),
    as.character(approx_n),
    if (is.null(top_mech)) "n/a" else paste(paste0(top_mech$aso_id, "=", top_mech$mechanism_label), collapse = "; ")
  ),
  stringsAsFactors = FALSE
), font_size = 10)
```

# ASO sequences and chemistry

```{r}
if (!is.null(qc)) {
  qc_show <- if (is_pdf) {
    qc %>%
      dplyr::transmute(
        aso = aso_id,
        sequence = cleaned_sequence,
        len = length_nt,
        GC = gc_percent,
        core = chemistry_core,
        arch = chemistry_architecture,
        gapmer = chemistry_gapmer,
        RNaseH = chemistry_rnase_h_direct_SHANK3_mRNA,
        m5dC = chemistry_five_methyl_dC
      )
  } else {
    qc %>% dplyr::select(aso_id, cleaned_sequence, length_nt, gc_percent,
                         chemistry_backbone, chemistry_moe, chemistry_five_methyl_dC,
                         chemistry_architecture, chemistry_rnase_h_compatible,
                         qc_warnings, status)
  }
  show_kable(qc_show, landscape = is_pdf, font_size = 9)
  tsv_note("results/01_sequence_qc.tsv")
}
```

**Chemistry rules applied throughout:** treat architecture as **likely_uniform_MOE**; gapmer **not supported** by current supplier notation; RNase-H cleavage of SHANK3 mRNA is **low_support** (favor steric/occupancy/RBP/structure); 5m-dC is **reported_but_needs_confirmation** (caveat only); FAM label is noted but **not** included in unlabeled mechanism interpretation; PS linkage pattern awaits supplier notation (map not invented). Canonical bases retained for alignment/motifs.

# Target gene and phenotype context

- Gene: **SHANK3** (neuronal synaptic scaffolding; dosage-sensitive ASD/Phelan-McDermid relevant gene).
- Cell types of interest: **neuron**, **neural precursor cell**.
- Phenotype for both ASOs: **increased SHANK3 expression**.
- Implication: mechanisms that **destroy** SHANK3 mRNA via RNase-H are directionally inconsistent with the stated phenotype for simple on-target knockdown (unless isoform switching/feedback is demonstrated).

# Transcript and genomic mapping (GRCh38)

```{r}
if (exact_n == 0) cat("**Ambiguity statement:** ", cfg$ambiguity_statement, "\n\n", sep = "")
if (!is.null(map)) {
  map_show <- if (is_pdf) {
    map %>%
      dplyr::transmute(
        aso = aso_id,
        match = shorten(match_class, 20),
        exact = exact_shank3_match,
        tx = shorten(transcript_id, 16),
        chrom = shorten(chrom, 12),
        pos = paste0(start, "-", end),
        str = strand,
        mm = mismatches,
        rank = rank
      )
  } else {
    map %>% dplyr::select(dplyr::any_of(c(
      "aso_id", "match_class", "exact_shank3_match", "approximate_shank3_match", "feature_type",
      "transcript_id", "chrom", "start", "end", "strand", "mismatches", "rank",
      "support_note", "status", "warning"
    )))
  }
  show_kable(map_show, landscape = is_pdf, font_size = 8)
  tsv_note("results/02_shank3_mapping.tsv")
}
```

Windows (+/- `r cfg$window_upstream` / `r cfg$window_downstream` nt) are written to `results/02_target_windows.fasta` and `.bed`.

# RNA structure and accessibility

ViennaRNA (`RNAfold`/`RNAplfold`) is invoked only via `system2()`. Missing tools produce warnings and empty numeric fields - **results are never fabricated**. All energies are canonical-RNA approximations unsuitable as exact MOE/PS/5m-dC biochemical measurements.

```{r}
if (!is.null(struct)) {
  struct_show <- if (is_pdf) {
    struct %>%
      dplyr::transmute(
        aso = aso_id,
        ctx = context,
        MFE = mfe_kcal_mol,
        site = shorten(site_structure_class, 16),
        unpaired = mean_unpaired_probability,
        hyb = approx_hybridization_energy,
        status = status
      )
  } else {
    struct %>% dplyr::select(dplyr::any_of(c(
      "aso_id", "window_id", "context", "mfe_kcal_mol", "site_structure_class",
      "mean_unpaired_probability", "approx_hybridization_energy",
      "status", "warning", "chemistry_caveat"
    )))
  }
  show_kable(struct_show, landscape = is_pdf, font_size = 8)
  tsv_note("results/03_rna_structure.tsv")
}
```

# RBP CLIP, motif, expression, and function evidence

Sources referenced: ENCODE eCLIP, UCSC ENCODE tracks, POSTAR3, oRNAment, ATtRACT, RBPDB, RBP2GO (see provenance). **No CLIP peaks are invented.**

## CLIP overlaps

```{r}
if (!is.null(clip)) {
  clip_show <- if (is_pdf) {
    clip %>%
      dplyr::transmute(
        aso = aso_id,
        rbp = ifelse(is.na(rbp), "-", as.character(rbp)),
        source = shorten(clip_source, 24),
        chrom = shorten(chrom, 10),
        site = paste0(site_start, "-", site_end),
        ov0 = overlap_0nt,
        strength = evidence_strength,
        status = status
      )
  } else {
    clip
  }
  show_kable(clip_show, landscape = is_pdf, font_size = 8)
  tsv_note("results/04_rbp_clip_overlap.tsv")
}
```

## Motif overlaps (sample)

```{r}
if (!is.null(motif)) {
  motif_show <- if (is_pdf) {
    utils::head(motif, 18) %>%
      dplyr::transmute(
        aso = aso_id,
        rbp = rbp,
        motif = shorten(motif_id, 14),
        start = match_start,
        end = match_end,
        strength = evidence_strength
      )
  } else {
    utils::head(motif, 30)
  }
  show_kable(motif_show, font_size = 9)
  tsv_note("results/04_rbp_motif_overlap.tsv")
}
```

## Expression priors (neuron / NPC)

```{r}
if (!is.null(expr)) {
  expr_show <- if (is_pdf) {
    utils::head(expr, 14) %>%
      dplyr::transmute(
        aso = aso_id,
        rbp = rbp,
        cell = shorten(cell_type, 18),
        expressed = expressed,
        status = status
      )
  } else {
    utils::head(expr, 20)
  }
  show_kable(expr_show, font_size = 9)
  tsv_note("results/04_rbp_expression.tsv")
}
```

## Function annotation

```{r}
if (!is.null(func)) {
  func_show <- if (is_pdf) {
    utils::head(func, 14) %>%
      dplyr::transmute(
        aso = aso_id,
        rbp = rbp,
        process = shorten(rna_process, 14),
        direction = shorten(regulatory_direction_hint, 18),
        status = status
      )
  } else {
    utils::head(func, 20)
  }
  show_kable(func_show, font_size = 9)
  tsv_note("results/04_rbp_function.tsv")
}
```

# Ranked RBP hypotheses

Transparent scoring: **+3 / +2 / +1 / -2** rules in `R/06_mechanism_ranking.R`. Top 3-5 RBPs per ASO attempt to include repressive, activating/stabilizing, and structure-related classes.

```{r}
if (!is.null(ranked)) {
  ranked_show <- ranked %>% dplyr::filter(in_top_report)
  if (is_pdf) {
    ranked_show <- ranked_show %>%
      dplyr::transmute(
        aso = aso_id,
        rank = rank,
        rbp = rbp,
        score = score,
        class = shorten(rbp_class, 16),
        direction = shorten(regulatory_direction_hint, 18)
      )
  } else {
    ranked_show <- ranked_show %>%
      dplyr::select(aso_id, rank, rbp, score, rbp_class, regulatory_direction_hint, score_rationale, warning)
  }
  show_kable(ranked_show, font_size = 9)
  tsv_note("results/06_ranked_rbps.tsv")
}
```

# TF and chromatin (conditional)

TF motifs are evaluated on **genomic DNA only** when locus class suggests nuclear regulatory / intronic / promoter / enhancer / antisense context. This evidence stream is separate from RBP analysis.

```{r}
if (!is.null(tf)) {
  tf_show <- if (is_pdf) {
    tf %>%
      dplyr::transmute(
        aso = aso_id,
        tf = ifelse(is.na(tf), "-", as.character(tf)),
        status = status,
        warning = shorten(warning, 40)
      )
  } else {
    tf
  }
  show_kable(tf_show, font_size = 9)
  tsv_note("results/05_tf_chromatin.tsv")
}
```

# Chemistry-aware mechanism ranking

Chemistry/mechanism rules apply the revised supplier-informed chemistry (likely_uniform_MOE; gapmer not supported; RNase-H direct SHANK3 mRNA low_support; PS protein engagement with linkage map pending; MOE steric/occupancy bias; 5m-dC reported-unconfirmed caveat; FAM excluded from unlabeled mechanism; structure hypothetical; sequence != proof). Labels scored as hypotheses only.

```{r}
if (!is.null(hyp)) {
  hyp_show <- hyp %>% dplyr::filter(rank <= 5)
  if (is_pdf) {
    hyp_show <- hyp_show %>%
      dplyr::transmute(
        aso = aso_id,
        rank = rank,
        mechanism = shorten(mechanism_label, 26),
        score = hypothesis_score,
        RNaseH = rnase_h_status,
        top_rbps = shorten(top_rbps, 28)
      )
  } else {
    hyp_show <- hyp_show %>%
      dplyr::select(aso_id, rank, mechanism_label, hypothesis_score, rnase_h_status,
                    exact_shank3_match, top_rbps, rationale, caveat)
  }
  show_kable(hyp_show, landscape = is_pdf, font_size = 8)
  tsv_note("results/07_mechanism_hypotheses.tsv")
}
```

# Per-ASO interpretation

```{r}
if (!is.null(hyp) && !is.null(ranked)) {
  for (aso in cfg$asos) {
    id <- aso$id
    cat("## ", id, " (`", aso$sequence, "`)\n\n", sep = "")
    tm <- hyp %>% dplyr::filter(aso_id == id, is_top)
    tr <- ranked %>% dplyr::filter(aso_id == id, in_top_report) %>% dplyr::arrange(dplyr::desc(score))
    cat("- Top mechanism hypothesis: **", tm$mechanism_label[1] %||% "unresolved", "**\n", sep = "")
    cat("- Top RBPs: ", paste(tr$rbp, collapse = ", "), "\n", sep = "")
    cat("- RNase-H: **low_support** (gapmer not supported; likely_uniform_MOE favors steric/occupancy; phenotype argues against simple on-target RNase-H knockdown)\n")
    cat("- Exact SHANK3 match present: ", any(map$aso_id == id & map$exact_shank3_match, na.rm = TRUE), "\n\n", sep = "")
  }
}
```

# Off-target considerations

A limited **chr22 genomic exact reverse-complement** screen is recorded in `results/02_offtargets_chr22.tsv`. This is **not** a full transcriptome off-target proof. Off-target mechanisms remain possible, especially without exact SHANK3 transcript confirmation.

# Validation recommendations

For every high-priority RBP, the following ten experiment types are recommended (each states which hypothesis it tests):

```{r}
if (!is.null(val)) {
  val_show <- if (is_pdf) {
    val %>%
      dplyr::group_by(aso_id, rbp) %>%
      dplyr::slice_head(n = 2) %>%
      dplyr::ungroup() %>%
      dplyr::transmute(
        aso = aso_id,
        rbp = rbp,
        experiment = shorten(experiment, 36),
        tests = shorten(hypothesis_tested, 42)
      )
  } else {
    utils::head(val, 60)
  }
  show_kable(val_show, landscape = is_pdf, font_size = 8)
  tsv_note("results/07_validation_recommendations.tsv")
}
```

# Resource provenance

```{r}
if (!is.null(prov)) {
  prov_show <- if (is_pdf) {
    prov %>%
      dplyr::transmute(
        resource = shorten(resource, 18),
        version = shorten(version, 22),
        build = genome_build,
        date = download_date
      )
  } else {
    prov
  }
  show_kable(prov_show, font_size = 9)
  tsv_note("data/raw/PROVENANCE.tsv")
} else {
  cat("No provenance table yet.\n")
}
```

# Limitations and missing information

- Architecture interpreted as **likely_uniform_MOE**; gapmer **not_supported_by_current_supplier_notation**.
- **RNase-H direct SHANK3 mRNA: low_support** — not assumed; steric/occupancy/RBP/structure favored.
- **5m-dC: reported_but_needs_confirmation** — caveat only; not treated as experimentally confirmed.
- **PS linkage pattern** set to `parse_from_supplier_notation`, but the supplier notation string is **not yet provided** (TODO; PS map not invented).
- **FAM label** present in supplier construct but **excluded** from unlabeled mechanism interpretation.
- Neuron/NPC **eCLIP peak overlaps** at the precise site generally unavailable in this automated run (metadata may be cached; peaks not fabricated).
- Motif/expression layers include **explicitly labeled priors**.
- Structure values omit MOE/PS/5m-dC parameterization.
- No experimental ASO engagement assay integrated.
- Genome builds other than GRCh38 were **not** used.
- PDF tables show **readable column subsets**; full-width result tables remain in `results/*.tsv`.

# Conclusions (non-confirmatory)

```{r}
cat("Both ASOs should be interpreted under chemistry-aware occupancy / RBP / structure / splicing hypotheses. Gapmer/RNase-H cleavage of SHANK3 mRNA has low support under likely_uniform_MOE. ")
cat("The highest-scoring mechanism labels are **hypotheses only**. ")
if (exact_n == 0) cat(cfg$ambiguity_statement, " ")
cat("Definitive mechanism assignment requires the validation suite above, architecture disclosure, and locus-resolved CLIP/structure/functional genetics in neuron/NPC systems.\n")
```

# Reproducibility

```r
cd shank3-aso-mechanism
Rscript -e 'targets::tar_make()'
```

Pipeline orchestrated with the `targets` package; modules in `R/01_*.R`-`R/07_*.R`; config in `_config.yml`.
