---
title: "SHANK3 ASO Target Mapping Report"
output:
  html_document:
    toc: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
root <- if (file.exists("_config.yml")) getwd() else if (file.exists("../_config.yml")) normalizePath("..") else getwd()
if (file.exists(file.path(root, "R/00_bootstrap.R"))) source(file.path(root, "R/00_bootstrap.R"))
cfg <- yaml::read_yaml(file.path(root, "_config.yml"))
map <- readr::read_tsv(file.path(root, "results/02_shank3_mapping.tsv"), show_col_types = FALSE)
prov <- tryCatch(readr::read_tsv(file.path(root, "data/processed/02_annotation_provenance.tsv"), show_col_types = FALSE), error = function(e) NULL)
```

# Scope

Genome build **GRCh38 only**. Gene: **SHANK3**. Mapping uses reverse-complement exact matches against SHANK3 transcripts and genomic locus; approximate 1-2 mismatch matches are secondary.

# Annotation provenance

```{r}
if (!is.null(prov)) knitr::kable(prov) else cat("Provenance file not found.")
```

# Mapping results

```{r}
knitr::kable(map)
```

# Ambiguity handling

```{r}
exact <- sum(map$exact_shank3_match, na.rm = TRUE)
if (exact == 0) {
  cat(cfg$ambiguity_statement)
} else {
  cat("Exact SHANK3 match rows:", exact, "\nNo invented transcripts were used.")
}
```

# Caveats

- Exact sequence match ≠ confirmed functional target engagement.
- Off-target screen in this pipeline is limited (chr22 genomic exact RC screen).
- Chemistry (PS/MOE/5m-dC) is not used to invent mapping coordinates.
